---
title: "Princeton Winter Session 2022: Introduction to Data Analysis in R"
output: html_notebook
---

# Data Types in R
## Vectors
```{r}
v1 = c(1,2,3,4,5)
v1
```

```{r}
v2 = c("a", "b", "c", c(3,2,1))
v2
```

## Environments in R
Listing objects for a given environment:
```{r}
ls()
```

Removing objects from the environment:
```{r}
rm("v2")
ls()
```

## Variables
Variables require an assignment operator:
```{r}
x=10
x+5
x
```

### Arithmetic
Vector operations:
```{r}
v1*x
```

```{r}
y = 2 + 3*v1
y
```

Finding the length of a vector:
```{r}
length(y)
```

Finding the maximum value in a vector:
```{r}
max(y)
```

Finding the minimum value in a vector:
```{r}
min(y)
```

Finding the arithmetic mean for the values in a vector:
```{r}
mean(y)
```

Exercises:

```{r}
nums = c(1,2,3,4,5)
nums_product = nums * 3 + 5
nums_product
```

```{r}
squares = nums^2
squares
```

```{r}
session_objects = ls()
length(session_objects)
```

Object introspection:

```{r}
class(nums)
```

```{r}
class("a")
```

## Boolean data types
```{r}
logic1=c(TRUE, TRUE, FALSE, FALSE)
logic1
```

```{r}
class(logic1)
```

Expressions can evaluate into boolean vector results:
```{r}
logic2 = v1 > 2
logic2
```

Finding the summation for all elements within a vector:
```{r}
sum(logic2)
```

```{r}
mean(logic2)
```

```{r}
sum(v1>2)
```

## Exercises

```{r}
cursor = ls()
class(cursor)
```

```{r}
max1 = max(v1)
class(max1)
```

```{r}
max2 = max(logic1)
class(max2)
```

```{r}
big_nums = nums < mean(nums)
big_nums
```

## Elements in vectors

Elements can be referenced by their indices:
```{r}
nums[1]
```
```{r}
nums[5]
```

Indices out of range for a vector will return a `NA` value:
```{r}
nums[6]
```

## Subsets of vectors

Vectors containing a subset of specific elements within a vector can be created:
```{r}
nums[1:4]
nums[c(1,2,3)]
nums[c(3,3,3)]
```

# Data Frames

## Data sets in R
Displaying the data sets available in R:
```{r}
data()
```

Loading a data set into the global scope:
```{r}
data(precip)
precip
```

```{r}
class(precip)
```

## Creating data frames
Creating a data frame from the data set:

```{r}
names(precip)
# Create the first column "city" with the names column and row values populating this
df = data.frame(city=names(precip), value=precip)
```

String representation of the data frame:

```{r}
str(df)
```

Rendering a table interface for the data frame:
```{r}
View(df)
```

## Selecting elements within a data frame
Selecting the first row and first column
```{r}
df[1,1]
```

Selecting the first row and all columns
```{r}
df[14,]
```

Selecting a subset of rows and all columns:
```{r}
df[2:6,]
```

Selecting the first six rows of the data frame:

```{r}
head(df)
```

Selecting the last six rows of the data frame:

```{r}
tail(df)
```

Accessing all row values within a specified column:

```{r}
df$city
```

Filtering the rows by the value in a single column:
```{r}
df[df$city == "Portland",]
```

Filtering the rows by an arithmetic expression:
```{r}
df[df$value > 50,]
```

# Plotting Visualizations

Generating a scatterplot for the row values within a column:
```{r}
plot(df$value)
```

Generating a line plot for the row values:
```{r}
plot(df$value, type='l')
```

Generating a histogram plot for the row values:
```{r}
hist(df$value)
```

Data frame attributes:

Inspecting the attributes of a data frame:

```{r}
?data.frame
```

Retrieving the column names for a data frame:

```{r}
colnames(df)
```

Casting variables into different data types

```{r}
s = "Princeton"
class(s)
```

For invalid values (e. g. casting `characters` to `integers`) only a `NA` is returned:
```{r}
as.numeric(s)
s2 = "123"
as.numeric(s2)
```
